---
title: "Incarceration Trends in the U.S. by Urbanicity"
output: html_document
---

### Introduction

- The problem domain of this dataset that I chose to study and analyze is the relationship between urbanicity and incarceration rates in various counties and states across the United States from 1990 onwards. Urbanicity refers to the classification of regions as rural, suburban, small/mid, or urban, and I'm interested in exploring how this factor might influence the rates of jail and prison populations. My analysis calculates the average total jail and prison population rate by urbanicity, identifies the urbanicity with the highest and lowest total prison population rates, and analyzes the trend of prison population rates across all years, grouped by urbanicity. By examining these selected variables, this study aims to reveal the underlying patterns and potential social implications of the relationship between urbanicity and incarceration. It may help shed light on the policies and social dynamics that lead to differences in incarceration rates across different urban classifications.


### Summary Information

```{r setup, include=FALSE}
source("summary.R")
knitr::opts_chunk$set(echo = TRUE)
```

- The analysis of incarceration trends across different urbanicities in the United States reveals distinct patterns. The urbanicity with the lowest prison population rate is rural, while the highest is found in small/mid areas. When examining the change in prison population rates over time, rural areas saw an increase of 1004.22 individuals, small/mid areas increased by 1080.09 individuals, suburban areas increased by 505.82 individuals, and urban areas increased by 124.3 individuals. These changes in population rates might reflect different policies, social dynamics, or economic factors affecting urban and rural areas differently. The findings shed light on the relationship between urbanicity and incarceration rates, offering insights that may guide policy and interventions.
 
### The Dataset

Who collected the data?  

- The Vera Institute of Justice collected the data. 

How was the data collected or generated?  

- The Vera organization collected data on U.S. incarceration to analyze patterns, particularly relating to race. Drawing from various sources such as the Bureau of Justice Statistics (BJS), the dataset offers a detailed county-level view. The goal is to fill gaps in prison population information, enabling more insightful and grounded studies.

Why was the data collected?  

- The data on racial disparities in incarceration was collected to analyze patterns of incarceration in the U.S. Existing information on prison populations was often insufficient for detailed studies, so this data, with its county-level variability, allows for more precise analysis. The aim is to identify systemic racial biases within the legal system, highlight areas of discrimination, and promote fairness and accountability. By understanding these contours and trends, law enforcement agencies can foster public awareness of systemic racism and work towards human rights within the legal framework.

How many observations (rows) are in your data? 

- There are 39 rows in the dataset. 

How many features (columns) are in the data?  

- There are 23 columns in the dataset. 

What, if any, ethical questions or questions of power do you need to consider when working with this data?  

- Analyzing the data related to racial disparities in incarceration raises significant ethical questions, particularly concerning the lack of context surrounding these observations and counts. The accuracy and proper representation of population sizes in prisons, especially in relation to race, are unclear. Furthermore, the data's focus on mere population counts may overlook essential aspects such as the treatment, crimes, or personal situations of those represented, reducing individuals to mere numbers.

- These considerations prompt questions about personal or sensitive information that might be traceable to individuals and concerns about ownership, access, and potential misuse of the data. Who collected it, who can access it, and who "owns" it can deeply influence how it is used, and for what purposes. Misinterpretation or misuse might further marginalize or stereotype certain racial or social groups. Thus, while working with this data, we must recognize that it represents actual human beings and be mindful of the conclusions we draw, as they can have real implications on lives and society's perception of different communities.

What are possible limitations or problems with this data?   (at least 200 words)

- The data related to U.S. incarceration has significant limitations, notably the absence of information in numerous states over extended periods, undermining analyses like racial comparisons. This lack of data raises questions about why certain information might be missing and weakens the conclusions that can be drawn.

- Additionally, the data sets lose context and human stories behind the numbers, potentially concealing disparities in treatment or unjust imprisonments. Temporal constraints, geographical inconsistencies, and a lack of distinction between crime types further hinder understanding. These combined limitations highlight the need for caution in interpreting this data and reveal the necessity for more comprehensive and context-aware analyses.

### Trends Over Time Chart
```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("chart1_example.R")
```

- The time trend chart illustrates the prison population rates across different levels of urbanity, revealing intriguing dynamics. Rural areas are seen to have the highest prison population rate, followed by small/mid-sized locations, suburban areas, and finally urban regions. This pattern diverges from common perceptions that associate higher incarceration rates with more urbanized locales. Over the examined period, the urban areas exhibit the most significant decline in prison population rates, while rural areas register the largest increase. These trends might reflect shifts in law enforcement practices, economic conditions, or societal changes unique to each level of urbanity. The unexpected rise in rural areas and the substantial drop in urban zones call for a deeper examination to understand the underlying causes and implications of these evolving patterns.

### Variable Comparison Chart

```{r, echo=FALSE, warning = FALSE,  code = readLines("chart2_example.R")}
```

- The scatter plot showing Total Jail Population Rate vs. Total Prison Population Rate uncovers a positive correlation between these variables, as evidenced by the upward trend line. A high concentration of data points in the 0-250 range indicates that most regions have relatively low incarceration rates. However, as the data points move towards the upper right corner, they become more dispersed. This dispersion reflects a less uniform pattern among areas with higher incarceration rates, suggesting variations in local policy or unique conditions. This plot offers a valuable insight into the relationship between jail and prison populations, hinting at underlying trends that may require further investigation.


### Map
```{r, message= FALSE, echo=FALSE}
eval(parse(text = readLines("chart3.R")))
```

- In the mapping of incarceration rates across various counties in Washington, a notable pattern emerges with the highest concentrations in Yakima and Ferry counties, which are characterized as rural areas. Conversely, more urbanized regions like the Puget Sound display a lighter color, indicating lower incarceration rates. These counties are depicted with contrasting shades on the map, allowing for an immediate identification of areas with differing levels of imprisonment. The stark difference between the rural and urban areas emphasizes a broader trend that seems to link rural counties with higher levels of incarceration while urban areas appear to have relatively lower rates. Whether driven by socio-economic factors, policies, or other variables, this pattern prompts further investigation into the underlying reasons for such disparate distributions between urban and rural regions, raising questions about the justice system and its interactions with different community environments.



